---
title: "Reproducible Vignette for Manuscript"
output: html_document
date: "2025-03-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
library(rstan)
library(reshape2)
seed = 12125
setwd("U:/Documents/repos/uncertainty_quantification/")
## set up the directories to where these files are stored locally 
model_dir <- paste0(getwd(),"/R/model/")
results_dir <- paste0(getwd(),"/R/model/samples/pcbs_2019/2024/gaza/")

source("R/0_setup.R")
```

# Model Estimation 

The "run_model" files contain the code used to obtain posterior sample estimates for each of the three age distribution sources. For brevity, we will use the GMoH age-distributions as an illustration in this vignette. 

```{r define_vars}

## get the sex-specific age distributions 
pi_x_moh <- readRDS("data/pi_x_moh_2024.rds")
pi_x_moh <- pi_x_moh[pi_x_moh$sex!="t",]

## reshape the age distributions for the shape that we need for the model 
pi_x= spread(pi_x_moh[,c("sex", "age", "pi_x_mean")], key=age, value=pi_x_mean) ## means of the age distributions
pi_sds= spread(pi_x_moh[,c("sex", "age", "pi_x_sd")], key=age, value=pi_x_sd) ## uncertainty with the age distributions 

## for the model, we we want E(log(theta)) and sd(log(theta)), so apply Delta method to the means and sds: 
pi_mu = log(pi_x[,-1])
pi_sd = pi_sds[,-1]/pi_x[,-1]

## read in exposure data:
master_forecast_dt <- readRDS("R/lc/data_plus_forecasts_v2.rds")
pcbs_exp  <- master_forecast_dt[master_forecast_dt$region=="Gaza Strip"&master_forecast_dt$year==2024&master_forecast_dt$sex%in%c("m", "f")&master_forecast_dt$source=="pcbs",]

## number of exposures by age
E_x = spread(pcbs_exp[,c("sex", "age","pop")], key=age, value=pop)
## exposures by age 
E_age =colSums(E_x[,-1])
## get total exposures 
E = sum(rowSums(E_x[,-1]))

## reshape the forecasted baseline mortality as well 
pcbs_mx<-  master_forecast_dt[master_forecast_dt$region=="Gaza Strip"&master_forecast_dt$year==2024&master_forecast_dt$sex%in%c("m", "f")&master_forecast_dt$source=="lc_pcbs_2022",]
D_x_pcbs= spread(pcbs_mx[,c("sex", "age","mx_noc")], key=age, value=mx_noc)

## age-sex specific mortality rates 
mu_x_pcbs <-  (D_x_pcbs[,-1])/E_x[,-1] 
## age specific mortality 
mu_age_pcbs <- colSums(D_x_pcbs[,-1])/E_age

### set the reported death toll: for Gaza 2024, this number is 23719
R =23719

## total number of sexes 
S = nrow(mu_x_pcbs)
## total number of age groups 
X = ncol(mu_x_pcbs)
## age groups 
x <- as.numeric(colnames(mu_x_pcbs))
```

Now that we have defined our variables and formatted them into the shapes needed for the model, we can compile and run the model. For brevity, we're running just 1 chain with 100 iterations and 50 burn in, but for the actual analysis, we did 4 chains with 2,000 iterations/1,000 burn in each. 

```{r run_bayesian_model, echo=FALSE}
## compile the model 
compiled_model <- stan_model(paste0(model_dir, "bmmr_coverage_intervals.stan"))

## run the model
model_out <- sampling(compiled_model,
                      iter =100,
                      warmup=50, #BURN IN
                      chains =1,
                      seed = seed,
                      control = list(max_treedepth = 60,
                                     adapt_delta=0.85),
                      data = list(
                        mu_x_noc = mu_x_pcbs,
                        mu_age_noc = mu_age_pcbs,
                        E_x = E_x[,-1],
                        E_age = E_age,
                        pi_x_hat = pi_mu, 
                        pi_sd = pi_sd, 
                        R = R,
                        S = S,
                        X= X))
```

# Compute Life Expectancies 

The "get_le0" file contains the code that takes in the posterior samples and outputs lifetable estimates for each MCMC iteration. We'll illustrate with just ten random samples, but for the analysis, we used 1,000 random samples from each model. 

```{r extract_samples}
## read in samples ()
file_names <- c("moh_samples_1", "moh_samples_2", "moh_samples_3", "moh_samples_4")
model_out <- read_stan_csv(paste0(results_dir, file_names,".csv"))

### extract the model-generated mortality distributions (incl WPP deaths)
mu_x_total = rstan::extract(model_out, pars=c("mu_x_total"))$mu_x_total
## mortality over sexes (only age specific )
mu_age_total =  rstan::extract(model_out, pars=c("mu_age_total"))$mu_age_total

## this is a 4000 (iterations) x 2 (sex) x 18 (age groups) array 
num_iter <- dim(mu_x_total)[1] ## grab the number of MCMC iterations 

## get 10 random samples of the model: 
set.seed(233)
ran_iter <- sample(c(1:num_iter), 10, replace=FALSE)

## this should now be an array with dimension ran_iter x 2 x 17 
mu_x_subset <- mu_x_total[ran_iter,,]
mu_age_subset <- mu_age_total[ran_iter,]
## get estimated life expectancies from the life table calculations 
```

As an illustration, we'll just get 10 lifetable calculations for females: 
```{r female_lifetables}
lifetable_f  <- list()
for(i in 1:10){
  lifetable_f[[i]] <- lifetable.mx(x, t(mu_x_subset[i, 1,]), sex="f")  # get life expectancy for women 
}

## get life expectancy tables into a dataset 
all_lifetable_f <- Reduce(rbind,lifetable_f)
```

Once we have the life. For illustration, here is the code to obtain the life expectancies for females: 
```{r compute_le0}
lifetable_f_age0 <- get_le0_dt(all_lifetable_f, "Females", 2024,  "GMoH Report", le0=c("", "", ""))
```


# Visualizations 
```{r pressure, echo=FALSE}

```